---
title: 'Rasch model with item bias'
subtitle: 'item bias'
author: 'dacarras'
date: '`r format(Sys.time(), "%a %b %d %X %Y")`'
output: github_document
---

# Case Description

- Dos cursos de matemáticas de octavo básico, presentan diferencias importantes en sus resultados en una guia de trabajo. Estas guías, son aplicadas como preparación para las evaluación de cierre de semestre. Son versiones muy cortas de cómo lucen las pruebas finales.
- El contenido de la guía aplicada incluía resolución de problemas (5 items), ejercicios de geometría (4 items), y 4 preguntas de algebra.
- Ambos cursos son enseñados por el mismo profesor, pero en momentos diferentes de la semana. Un curso tiene clases los Lunes, y el otro curso tiene clases los Viernes. Sin embargo, las pruebas de cierre de trimestre y sus ensayos, son realizadas de forma sistemática los Miércoles.
- Las diferencias entre las notas obtenidas por ambos cursos desaparecen, cuando las preguntas de algebra son ignoradas.
- **¿Quizás el curso del Viernes, no alcanzo a revisar los contenidos de algebra?**


# Cargar datos

## Abrir datos

```{r , echo=TRUE}

#--------------------------------------------------------------------
# load data
#--------------------------------------------------------------------

data_wide <- readRDS('rasch_bias.rds')

# ----------------------------------------------- 
# check data format
# -----------------------------------------------

library(dplyr)
r4sda::variables_table(data_wide) %>%
knitr::kable()

```

## Re-structure data

```{r , echo=TRUE}

#--------------------------------------------------------------------
# prepare data
#--------------------------------------------------------------------

# ----------------------------------------------- 
# variables names
# -----------------------------------------------

data_stack <- data_wide %>%
              mutate(id_i = seq(1:nrow(.))) %>%
              tidyr::gather('item', 'resp', -u, -id_i) %>%
              dplyr::glimpse()

```


# Item bias with the logistic method

## Item bias with `difR` with logistic method

```{r , echo=TRUE}

#--------------------------------------------------------------------
# item bias
#--------------------------------------------------------------------

# ----------------------------------------------- 
# display results
# -----------------------------------------------

difR::difLogistic(
  Data = data_wide, 
  group = 'u', 
  focal.name = 0, 
  alpha = .05, 
  type = 'udif', 
  criterion = 'LRT'
  )

# Note: focal group is Friday, the gruoup without
#       without the oportunity to learn.


# ----------------------------------------------- 
# create dif object
# -----------------------------------------------

dif_results <- difR::difLogistic(
  Data = data_wide, 
  group = 'u', 
  focal.name = 0, 
  alpha = .05, 
  type = 'udif', 
  criterion = 'LRT'
  )

# ----------------------------------------------- 
# plot per item
# -----------------------------------------------

# problem items
plot(dif_results, plot = "itemCurve", item = 1, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 2, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 3, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 4, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 5, itemFit = "null")

# geometry items
plot(dif_results, plot = "itemCurve", item = 6, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 7, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 8, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 9, itemFit = "null")

# algebra items
plot(dif_results, plot = "itemCurve", item = 10, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 11, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 12, itemFit = "null")
plot(dif_results, plot = "itemCurve", item = 13, itemFit = "null")


```

## Replicate statistics manually

```{r , echo=TRUE}

#--------------------------------------------------------------------
# item bias
#--------------------------------------------------------------------

# ----------------------------------------------- 
# separate response matrix
# -----------------------------------------------

items_data <- dplyr::select(data_wide, i01:i13)

# ----------------------------------------------- 
# create total score
# -----------------------------------------------

data_dif <- data_wide %>%
            mutate(score = r4sda::mean_score(items_data)) %>%
            mutate(z_score = r4sda::z_score(score))

#-------------------------------------------------
# fit logistic method for uniform dif manually
#-------------------------------------------------

m0 <- glm(i13 ~ z_score, data = data_dif, family = "binomial")
m1 <- glm(i13 ~ z_score + u, data = data_dif, family = "binomial")

# ----------------------------------------------- 
# display results
# -----------------------------------------------

texreg::screenreg(list(m0, m1), 
    type = 'un',
    star.symbol = "*", 
    center = TRUE, 
    doctype = FALSE,
    dcolumn = TRUE, 
    booktabs = TRUE,
    single.row = TRUE)

# ----------------------------------------------- 
# replicate lrt stat
# -----------------------------------------------

# get LRT for null model
d0 <- broom::glance(m0) %>%
dplyr::select(deviance) %>%
dplyr::pull()

# get LRT for model 1
d1 <- broom::glance(m1) %>%
dplyr::select(deviance) %>%
dplyr::pull()

# replicate stat
d0-d1

# ----------------------------------------------- 
# replicate r2 effect size
# -----------------------------------------------

DescTools::PseudoR2(m1, "Nagel") - DescTools::PseudoR2(m0, "Nagel")

```

## Item bias with anchors

```{r , echo=TRUE}

#--------------------------------------------------------------------
# item bias
#--------------------------------------------------------------------

# ----------------------------------------------- 
# display results
# -----------------------------------------------

difR::difLogistic(
  Data = data_wide, 
  group = 'u', 
  focal.name = 0, 
  alpha = .05, 
  type = 'udif', 
  criterion = 'LRT'
  )

# Note: focal group is Friday, the gruoup without
#       without the oportunity to learn.


# ----------------------------------------------- 
# create dif object
# -----------------------------------------------

dif_anchored <- difR::difLogistic(
  Data = data_wide, 
  anchor = 1:9,
  group = 'u', 
  focal.name = 0, 
  alpha = .05, 
  type = 'udif', 
  criterion = 'LRT'
  )

# ----------------------------------------------- 
# plot per item
# -----------------------------------------------

# algebra items
plot(dif_anchored, plot = "itemCurve", item = 10, itemFit = "null")
plot(dif_anchored, plot = "itemCurve", item = 11, itemFit = "null")
plot(dif_anchored, plot = "itemCurve", item = 12, itemFit = "null")
plot(dif_anchored, plot = "itemCurve", item = 13, itemFit = "null")

```



# Item bias as an special case of explanatory response model problem

## Fit Rasch model as a mixed model

$$ln[Pr(y_{ip} = 1)] = \theta_{.p} - \delta_{1-13}item$$

```{r , echo=TRUE}

#--------------------------------------------------------------------
# fit rasch model
#--------------------------------------------------------------------

glm_00 <- lme4::glmer(resp ~ -1 + item + (1|id_i), data = data_stack, family = binomial('logit'), nAGQ = 15)

# ----------------------------------------------- 
# display results
# -----------------------------------------------

broom.mixed::tidy(glm_00) %>%
knitr::kable(., digits = 2)

```


# Item bias as mixed model


## Fit Rasch model as a mixed model

$$ln[Pr(y_{ip} = 1)] = \theta_{.p} - \delta_{1-13}item$$

```{r , echo=TRUE}

#--------------------------------------------------------------------
# fit rasch model
#--------------------------------------------------------------------

glm_00 <- lme4::glmer(resp ~ -1 + item + (1|id_i), data = data_stack, family = binomial('logit'), nAGQ = 15)

# ----------------------------------------------- 
# display results
# -----------------------------------------------

broom.mixed::tidy(glm_00) %>%
knitr::kable(., digits = 2)

```

## Fit Rasch model as a latent regression

$$ln[Pr(y_{ip} = 1)] = \theta_{.p} - \delta_{1-13}item + \beta_{.p}u_{.p}$$

```{r , echo=TRUE}
#--------------------------------------------------------------------
# latent regression main effect
#--------------------------------------------------------------------

glm_01 <- lme4::glmer(resp ~ -1 + item + u + (1|id_i), data = data_stack, family = binomial('logit'), nAGQ = 15)

# ----------------------------------------------- 
# display results
# -----------------------------------------------

broom.mixed::tidy(glm_01) %>%
knitr::kable(., digits = 2)

```

## Fit Rasch model as a latent regression including the DIF effects

$$ln[Pr(y_{ip} = 1)] = \theta_{.p} - \delta_{1-13}item + \beta_{.p}u_{.p} + \gamma_{ip}(item_{i.}u_{.p})$$

```{r , echo=TRUE}
#--------------------------------------------------------------------
# model with bias
#--------------------------------------------------------------------

glm_02 <- lme4::glmer(resp ~ -1 + item + u + item:u + (1|id_i), data = data_stack, family = binomial('logit'), nAGQ = 15)

# ----------------------------------------------- 
# display results
# -----------------------------------------------

broom.mixed::tidy(glm_02) %>%
knitr::kable(., digits = 2)

```

## Compare fitted models

```{r , echo=TRUE}
#--------------------------------------------------------------------
# model comparison
#--------------------------------------------------------------------

# ----------------------------------------------- 
# variables names
# -----------------------------------------------

texreg::screenreg(list(glm_00, glm_01, glm_02), 
  star.symbol = "\\*",
  single.row = TRUE, 
  center = TRUE, 
  doctype = FALSE,
  dcolumn = TRUE, 
  booktabs = TRUE)


```


## Commented output

```text

=================================================================================
                       Model 1             Model 2             Model 3           
---------------------------------------------------------------------------------
itemi01                   1.69 (0.35) ***     1.47 (0.36) ***     1.80 (0.41) *** # these are deltas
itemi02                   1.51 (0.34) ***     1.29 (0.35) ***     1.34 (0.38) ***
itemi03                   0.94 (0.32) **      0.72 (0.33) *       0.83 (0.36) *  
itemi04                   1.42 (0.34) ***     1.20 (0.35) ***     1.24 (0.38) ** 
itemi05                   0.57 (0.31)         0.35 (0.32)         0.55 (0.36)    
itemi06                   0.29 (0.31)         0.07 (0.32)         0.37 (0.36)    
itemi07                   0.22 (0.31)        -0.00 (0.32)         0.09 (0.35)    
itemi08                  -0.27 (0.31)        -0.49 (0.32)        -0.54 (0.36)    
itemi09                  -0.56 (0.31)        -0.77 (0.33) *      -0.64 (0.36)    
itemi10                  -0.34 (0.31)        -0.56 (0.32)        -0.84 (0.37) *  
itemi11                  -0.70 (0.32) *      -0.92 (0.33) **     -1.14 (0.38) ** 
itemi12                  -2.33 (0.40) ***    -2.55 (0.41) ***    -3.24 (0.56) ***
itemi13                  -1.50 (0.34) ***    -1.72 (0.36) ***    -2.43 (0.46) ***
u                                             0.97 (0.42) *      -0.48 (0.81)     # <- measure of impact
itemi02:u                                                         1.23 (1.05)     #    this is beta
itemi03:u                                                         0.91 (0.98)    
itemi04:u                                                         1.33 (1.04)    
itemi05:u                                                         0.53 (0.95)    
itemi06:u                                                         0.10 (0.94)    
itemi07:u                                                         0.99 (0.95)    
itemi08:u                                                         1.62 (0.95)    
itemi09:u                                                         0.81 (0.95)    
itemi10:u                                                         2.58 (0.98) **  # items with diff
itemi11:u                                                         2.23 (0.96) *   # these are gamma
itemi12:u                                                         3.10 (1.05) **  #
itemi13:u                                                         3.51 (1.00) *** #
---------------------------------------------------------------------------------
AIC                    1156.68             1153.37             1146.94           
BIC                    1225.94             1227.57             1280.50           
Log Likelihood         -564.34             -561.68             -546.47           
Num. obs.              1040                1040                1040              
Num. groups: id_i        80                  80                  80              
Var: id_i (Intercept)     2.05                1.89                2.07           
=================================================================================
*** p < 0.001, ** p < 0.01, * p < 0.05

```

>Note: when a mixed model is fitted, to retrieve a Rasch Model, the delta parameters are "reverse". The easier items are "positive", and the most difficult items are "negative" (i.e. preceded with a "-" sign).

## References

- De Boeck, P., Bakker, M., Zwitser, R., Nivard, M., Hofman, A., Tuerlinckx, F., & Partchev, I. (2011). The Estimation of Item Response Models with the lmer Function from the lme4 Package in R. Journal Of Statistical Software, 39(12), 1–28. https://doi.org/10.18637/jss.v039.i12


# Overall DIF detection

```{r , echo=TRUE, warning = FALSE, eval = FALSE}

#--------------------------------------------------------------------
# diff detection
#--------------------------------------------------------------------

# ----------------------------------------------- 
# no anchor
# -----------------------------------------------

# Note: DIF is assess without specifying comparable items.

difR::dichoDif(
  Data=data_wide, 
  group="u", 
  focal.name=1, 
  method=c("MH","LRT"),
  model="1PL",
  nrIter=20)



# ----------------------------------------------- 
# with anchor
# -----------------------------------------------

# Note: DIF is assess using items i01-i09 as comparable items

difR::dichoDif(
  Data=data_wide, 
  group="u", 
  focal.name=1, 
  method=c("MH","LRT"),
  model="1PL",
  anchor = 1:9,
  nrIter=20)


# ----------------------------------------------- 
# with purify
# -----------------------------------------------

# Note: DIF is assess letting the library searching for "comparable" items.

difR::dichoDif(
  Data=data_wide, 
  group="u", 
  focal.name=1, 
  method=c("MH","LRT"),
  model="1PL",
  purify=TRUE,
  nrIter=20)

# ----------------------------------------------- 
# shortened version
# -----------------------------------------------

# Note: DIF is assess only among the comparable items.

selected_data <- dplyr::select(raw_data, i01:i09, u)

difR::dichoDif(
  Data=selected_data, 
  group="u", 
  focal.name=1, 
  method=c("MH","LRT"),
  model="1PL",
  purify=TRUE,
  nrIter=20)


```

## References

Magis, D., Béland, S., Tuerlinckx, F., & de Boeck, P. (2010). A general framework and an R package for the detection of dichotomous differential item functioning. Behavior Research Methods, 42(3), 847–862. https://doi.org/10.3758/BRM.42.3.847

