---
title: "Thurstone's Box Problem"
subtitle: 'EFA & CFA'
author: 'dacarras'
output: github_document
---

# Case Description

- Empleamos la variante del Thurstone's Box Problem de Kaiser & Horst (1975).
- Los autores generan una base de datos de 10 indicadores, de cuarenta cajas diferentes, agregandoles error aleatorio a las medidas.
- Sobre esta matriz de datos, se espera ajustar un modelo con tres factores.


# Cargar datos

## Abrir datos

```{r , echo=TRUE}

#--------------------------------------------------------------------
# load data from url
#--------------------------------------------------------------------

box_data<-read.csv("https://www.evernote.com/shard/s42/sh/d0f20791-c8e5-4627-972e-5c76625f0ffb/695007b60806fd4ecfa7dbb73c67f8d1/res/4ac06b03-0a8e-4c3d-8d57-29f6e71a6598/box_prob.csv")

# ----------------------------------------------- 
# check data format
# -----------------------------------------------

library(dplyr)
r4sda::variables_table(box_data) %>%
knitr::kable()

# ----------------------------------------------- 
# isolate data from indicators
# -----------------------------------------------

box_items <- dplyr::select(box_data, i1:i10)


# ----------------------------------------------- 
# display first ten cases
# -----------------------------------------------

knitr::kable(box_items[1:10,], digits = 2)


```

# Dimensionalidad

## Scree test

```{r , echo=TRUE}

#--------------------------------------------------------------------
# dimensionality
#--------------------------------------------------------------------

# ----------------------------------------------- 
# load data
# -----------------------------------------------

box_data<-read.csv("https://www.evernote.com/shard/s42/sh/d0f20791-c8e5-4627-972e-5c76625f0ffb/695007b60806fd4ecfa7dbb73c67f8d1/res/4ac06b03-0a8e-4c3d-8d57-29f6e71a6598/box_prob.csv")

# ----------------------------------------------- 
# isolate data from indicators
# -----------------------------------------------

box_items <- dplyr::select(box_data, i1:i10)

# ----------------------------------------------- 
# scree test
# -----------------------------------------------

psych::scree(box_items, factors = FALSE, pc = TRUE, hline = 1)

# Nota: scree test generado sobre los eigenvalues de un 
#       análisis de componentes principales.

```

## Parallel

```{r , echo=TRUE}

#--------------------------------------------------------------------
# dimensionality
#--------------------------------------------------------------------

# ----------------------------------------------- 
# load data
# -----------------------------------------------

box_data<-read.csv("https://www.evernote.com/shard/s42/sh/d0f20791-c8e5-4627-972e-5c76625f0ffb/695007b60806fd4ecfa7dbb73c67f8d1/res/4ac06b03-0a8e-4c3d-8d57-29f6e71a6598/box_prob.csv")

# ----------------------------------------------- 
# isolate data from indicators
# -----------------------------------------------

box_items <- dplyr::select(box_data, i1:i10)

# ----------------------------------------------- 
# parallel
# -----------------------------------------------

psych::fa.parallel(box_items, fa = "fa", show.legend = FALSE, fm = "ml")

```

## MAP

```{r , echo=TRUE}

#--------------------------------------------------------------------
# dimensionality
#--------------------------------------------------------------------

# ----------------------------------------------- 
# load data
# -----------------------------------------------

box_data<-read.csv("https://www.evernote.com/shard/s42/sh/d0f20791-c8e5-4627-972e-5c76625f0ffb/695007b60806fd4ecfa7dbb73c67f8d1/res/4ac06b03-0a8e-4c3d-8d57-29f6e71a6598/box_prob.csv")

# ----------------------------------------------- 
# isolate data from indicators
# -----------------------------------------------

box_items <- dplyr::select(box_data, i1:i10)

# ----------------------------------------------- 
# traditional output
# -----------------------------------------------

psych::VSS(x=box_items,fm="mle", plot = FALSE)


# ----------------------------------------------- 
# output as plot
# -----------------------------------------------

map_out <- psych::VSS(box_items, rotate ="varimax", plot = FALSE)

# extract MAP values and create vector of factors
map_vector  <- data.frame(map=map_out$map)
map_factors <- data.frame(factors=c(1:nrow(map_vector)))

# create data for plot
map_data <- dplyr::bind_cols(map_factors, map_vector)

# get values
map_value    <- min(map_vector, na.rm = TRUE)
map_retained <- dplyr::filter(map_data, map == min(map_vector, na.rm=TRUE)) %>% 
                dplyr::select(factors) %>% 
                as.numeric()

# plot MAP results
plot(map_data, 
  main = 'MAP criteria', 
  ylab = 'MAP values', 
  xlab = 'Number of factors', 
  ylim = c(0,.20), 
  type = 'p', 
  col = 'black', 
  pch=16)
abline(
  h = map_value, 
  v = map_retained, 
  col = "red", 
  lty=2, 
  lwd=1)

```

## EFA

```{r , echo=TRUE}

#--------------------------------------------------------------------
# dimensionality
#--------------------------------------------------------------------

# ----------------------------------------------- 
# load data
# -----------------------------------------------

box_data<-read.csv("https://www.evernote.com/shard/s42/sh/d0f20791-c8e5-4627-972e-5c76625f0ffb/695007b60806fd4ecfa7dbb73c67f8d1/res/4ac06b03-0a8e-4c3d-8d57-29f6e71a6598/box_prob.csv")

# ----------------------------------------------- 
# isolate data from indicators
# -----------------------------------------------

box_items <- dplyr::select(box_data, i1:i10)

# ----------------------------------------------- 
# exploratory factor analysis
# -----------------------------------------------

library(psych)
library(GPArotation)
efa_fit <- fa(box_items, fm="mle", nfactors=3, rotate="varimax")

# ----------------------------------------------- 
# exploratory factor analysis
# -----------------------------------------------

efa_fit


```

## CFA

```{r , echo=TRUE}

#--------------------------------------------------------------------
# dimensionality
#--------------------------------------------------------------------

# ----------------------------------------------- 
# load data
# -----------------------------------------------

box_data<-read.csv("https://www.evernote.com/shard/s42/sh/d0f20791-c8e5-4627-972e-5c76625f0ffb/695007b60806fd4ecfa7dbb73c67f8d1/res/4ac06b03-0a8e-4c3d-8d57-29f6e71a6598/box_prob.csv")

# ----------------------------------------------- 
# isolate data from indicators
# -----------------------------------------------

box_items <- dplyr::select(box_data, i1:i10)

# ----------------------------------------------- 
# confirmatory factor analysis
# -----------------------------------------------

lavaan_model <- '

axis_x =~ start( 0.01)*i1 
axis_x =~ start( 0.94)*i2 
axis_x =~ start(-0.01)*i3 
axis_x =~ start( 0.90)*i4 
axis_x =~ start( 0.12)*i5 
axis_x =~ start( 0.41)*i6 
axis_x =~ start( 0.82)*i7 
axis_x =~ start( 0.42)*i8 
axis_x =~ start( 0.70)*i9 
axis_x =~ start( 0.83)*i10

axis_y =~ start(0.00)*i1 
axis_y =~ start(0.21)*i2 
axis_y =~ start(0.97)*i3 
axis_y =~ start(0.18)*i4 
axis_y =~ start(0.85)*i5 
axis_y =~ start(0.88)*i6 
axis_y =~ start(0.07)*i7 
axis_y =~ start(0.42)*i8 
axis_y =~ start(0.65)*i9 
axis_y =~ start(0.09)*i10


axis_z =~ start( 0.38)*i1 
axis_z =~ start(-0.18)*i2 
axis_z =~ start( 0.09)*i3 
axis_z =~ start( 0.23)*i4 
axis_z =~ start( 0.42)*i5 
axis_z =~ start(-0.04)*i6 
axis_z =~ start( 0.51)*i7 
axis_z =~ start( 0.78)*i8 
axis_z =~ start(-0.21)*i9 
axis_z =~ start( 0.34)*i10


axis_x ~~ 0*axis_y
axis_x ~~ 0*axis_z
axis_y ~~ 0*axis_z

'
# ----------------------------------------------- 
# fit lavaan model
# -----------------------------------------------

cfa_fit <- lavaan::sem(lavaan_model, 
            data = box_items,
            mimic='mplus',
            estimator = 'MLR')

# ----------------------------------------------- 
# display results
# -----------------------------------------------

lavaan::summary(cfa_fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)


```

## References

Kaiser, H. F., & Horst, P. (1975). A Score Matrix for Thurstone’s Box Problem. Multivariate Behavioral Research, 10(1), 17–26. http:// doi.org/10.1207/s15327906mbr1001_2


